Effects of bariatric surger and weight loss on resting-state functional connectivity

### 2020-11-29 draft_rsfmri.Rmd
update introduction

Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.

Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. On the other hand, dietary interventions such as caloric restriction and related metabolic improvements and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
Behavioral intervention require ... and often fail to yield the desired success regarding weight loss, especially in in morbid obesity [@]. There are alternative interventions at hand.
<!-- bars -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions and co-morbidities, such as diabetes. Roux-en-Y gastric bypass (RYGB) is the most common procedure, as it not only restricts the food intake by reducing stomach volume, but also affects the absorption of macronutrients and the hormonal profile [@]. Other common procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. 
<!-- Potential mechanisms? -->
Precise mechanisms are yet to be elucidated, but as a first step, a comprehensive understanding of the concurrent physiological changes underlying weight loss and recovery is needed to better infer potential mechanisms and support interventions proven to be effective. <!-- / immitate interventions that have been proven effective with less invasive alternatives -->

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in this population, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
<!-- mechanism reward and appetite, overating -->
Besides malabsorption and restricted food ingestion, patients report post-surgical changes in food preferences and eating behavior. Those psychophysiological and behavioral effects may be related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which a sustainable weight loss [@Mulla_2017, @Nudel_2019].

@@

Weight regulation and energy expenditure is dependent on a complex interplay of homeostatic signal and internal sensory processing. Obesity may be cause or effect of a decompensation of the psychophysiological system that ensures homeostasis.
The default mode network is commonly associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> It is mainly composed of the posterior cingulate cortex (PCC)/ Precuneus (PC), the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- difference lean obese --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2016]. Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
A longitudinal study that confirmed the stronger connectivity (of PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG). However, default mode connectivity in these regions seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable 1 year after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
This is consistent with other findings of reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC [@Li_2018] <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery.

@@

On the other hand, connectivity also served as a predictor of weight loss after surgery.
On the other hand, there are also indicators of reduced functional integrity, e.g. for the anterior cingulate cortex (ACC) at resting state compared to lean individuals [@Kullmann_2011].

@@

# distributions per group
# BMI 
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
tab_age <- describeBy(data=dfBL, x=dfBL$Age_BL, group=dfBL$condition, mat=T)
# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) + 
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555"),
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tabdescr, echo=FALSE}

>>>>

<!-- obesity and interventions -->
Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.
_something about comorbidities_
Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. 
Luckily, dietary interventions such as caloric restriction and related metabolic improvements, and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
In some cases behavioral intervention fail to yield the desired success regarding weight loss. There are non-behavioral alternatives at hand.
<!-- bars most invasive (brain metabolism), last resort -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions, and co-morbidities, such as diabetes. Roux-en-Y gastric bypass (RYGB) is the most common procedure as it not only restricts the food intake by reducing stomach volume but also affects the absorption of macronutrients and the hormonal profile [@]. <!-- ref for "common", ref for procedure -->
Other widely used procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. <!-- mehr zu Veränderung in Gasto-intestinaler Physiologie -->

<!-- Potential mechanisms? -->
Precise mechanisms are yet to be elucidated. As a first step, a comprehensive understanding of the concurrent physiological changes underlying weight loss and recovery is needed. This will inform inferences about potential mechanisms and either support interventions proven to be effective or ideally allow imitating them will less invasive alternatives.

Recently, research focuses have slowly shifted to investigate brain metabolism structural and functional changes after surgery.
<!-- severe nutritional stress [@Crujeiras_2017] -->

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in the obese population, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
<!-- mechanism reward and appetite, overating -->
Besides malabsorption and restricted food ingestion, patients report post-surgical changes in food preferences and eating behavior. Those psychophysiological and behavioral effects may be related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which a sustainable weight loss [@Mulla_2017, @Nudel_2019].

@@

Weight regulation and energy expenditure is dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. Obesity may be cause or effect of a decompensation of the psychophysiological system that ensures homeostasis.
The default mode network is commonly associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> It is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- difference lean obese --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2016]. Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
A longitudinal study that confirmed the stronger connectivity (of PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG). However, default mode connectivity in these regions seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable 1 year after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
This is consistent with other findings of reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC [@Li_2018] <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery.

@@

On the other hand, connectivity also served as a predictor of weight loss after surgery.

On the other hand, there are also indicators of reduced functional integrity, e.g. for the anterior cingulate cortex (ACC) at resting state compared to lean individuals [@Kullmann_2011].

@@

# distributions per group
# BMI 
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
tab_age <- describeBy(data=dfBL, x=dfBL$Age_BL, group=dfBL$condition, mat=T)
# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) + 
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tabdescr, echo=FALSE}
